Global Distribution of Unintended Pregnancies and Contraceptive Usage

### 2024-02-18 code/processing-code/processingfile-v1.qmd
Part 2 Updates
I found new variables to use as covariates for our updated research question. I added those variables to the raw data that is being processed, explored them, and created a new variable. I also made notes on how to start exploring the data.

```{r}
library(readr) # for loading csv files
library(readxl) # for loading excel files (codebook)
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(naniar) #to explore data missingness
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(tidymodels) #to impute missing data
library(recipes) #to impute missing data
```

@@

```{r}
# path to data using here() function
data_location <- here("data","raw-data","AIU All Women Dataset.csv")
rawdata <- read_csv(data_location)
```

@@

```{r}
# write file path to the codebook
codebook_location <- here("data", "AIU All Women Dataset Codebook.xlsx")

@@

```{r}
dplyr::glimpse(rawdata)
summary(rawdata)
head(rawdata)
skimr::skim(rawdata)
```
# Identifying variables to be cleaned

```{r}
rawdata2 <- rawdata %>% 
                select(country, region, subregion, pct_upreg, upreg_subreg, modmethods_3549, ster_3549, larc_3549, sarc_3549, tradmethods_3549, nomethod_3549)

skimr::skim(rawdata2)
```

@@

First, the `region` variable is a character variable, but it will be more useful later as a factor. We can easily mutate it from a character variable to factor variable. All the remaining variables are numeric as they should be. 
```{r}
# convert character variable to factor variable
rawdata2 <- rawdata2 %>% 
              mutate(region = as.factor(region))

@@

Second, there are 24 entries missing for the percentage of all unintended pregnancies by country `pct_upreg`, which accounts for 18.18% of the observations. After isolating the observations with this missing variable, we see that the majority of countries in this category are located in Asia. This does not make sense because the majority of countries in the raw dataset are located in Africa, which suggests that the data are not missing at random. 
```{r}
# find missing percentage for each variable
miss_var_summary(rawdata2)

@@

We should impute the missing data here to avoid introducing bias because 18.18% of the data is missing, and this variable is the best outcome variable to answer the research question. After finding the summary statistics of this variable, the median is 0.4701 and the mean is 0.4679, which suggests that there are minimal outliers in the data.  
```{r}
# find mean to use for imputed data
summary(rawdata2$pct_upreg)

@@

```
Once the missing values have been imputed, we should further clean the `pct_upreg` variable to have a typical percentage format. 
```{r}
rawdata2 <- rawdata2 %>% 
              mutate(pct_upreg = pct_upreg * 100)
```

We also need to evaluate the 108 missing entries of the percentage of unintended pregnancies by subregion `upreg_subreg`. Since 81.81% of this variable is missing, it does not make sense to include it in the analysis because we do not have enough information to impute the missing data. We will remove this variable for the sake of this analysis.
```{r}
# isolate the percentage of all unintended pregnancies by subregion to discover trends in missingness
variable_with_missing <- "upreg_subreg"
missing_obs2 <- rawdata2[is.na(rawdata2[[variable_with_missing]]), ]
other_variables2 <- colnames(rawdata2)[!colnames(rawdata2) == variable_with_missing]
pct_subregmissing <- missing_obs2[, other_variables2]
print(pct_subregmissing)

@@

Each of the variables about birth control methods, `mod_methods3549`, `ster_3549`, `larc_3549`, `sarc_3549`, `tradmethods_3549`, `nomethod_3549`, are not very intuitive, so they should be renamed. Since we are only studying one age group, 35 to 49 years, it is unnecessary for those numbers to be included in each variable name. It also makes sense to rename `pct_upreg` to be more descriptive since we already know it is percentage format. 
```{r}
# rename predictor variables
rawdata2 <- rawdata2 %>% rename(modern = modmethods_3549,
                                sterile = ster_3549, 
                                longactrev = larc_3549,
                                shortact = sarc_3549,
                                tradition = tradmethods_3549,
                                nomethod = nomethod_3549,
                                unint_preg = pct_upreg)
# check if renaming worked
head(rawdata2)
```
To answer our research question about how unintended pregnancies are related to birth control methods, it makes sense to have a variable with the total number of women using birth control in each country. This will service as the main predictor variable, and each type of birth control can serve as a covariate. 
```{r}
# create new variable with total number on birth control of any form
rawdata2 <- rawdata2 %>% 
              mutate(totalbc = modern + sterile + longactrev + shortact + tradition + nomethod)

# check new variable is created 
head(rawdata2)
```

@@

```{r}
processeddata <- rawdata2
```

@@

```{r}
save_data_location <- here("data","processed-data","processeddata.rds")
saveRDS(processeddata, file = save_data_location)
```

>>>>

```{r}
#| message: false
library(readr) # for loading csv files
library(readxl) # for loading excel files (codebook)
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(naniar) #to explore data missingness
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(tidymodels) #to impute missing data
library(recipes) #to impute missing data
```

@@

```{r}
#| message: false
# path to data using here() function
data_location <- here("data","raw-data","AIU All Women Dataset.csv")
rawdata <- read_csv(data_location)
```

@@

```{r}
#| message: false
# write file path to the codebook
codebook_location <- here("data", "AIU All Women Dataset Codebook.xlsx")

@@

```{r}
#| message: false
dplyr::glimpse(rawdata)
summary(rawdata)
head(rawdata)
skimr::skim(rawdata)
```
# Identifying variables to be cleaned 
We want to explore the relationship between unintended pregnancies and birth control methods across different age groups. We will use three demographic variable: `country`, `region`, and `subregion`. The variable for the total percentage of unintended pregnancies, `pct_upreg` will be the main outcome variable. We could also use `upreg_subreg` to explore unintended pregnancies by subregion. The predictor variables will include percentage of unintended pregnancies across three types of birth control: long acting reversible methods (`pct_upreg_larcster`), short term reversible methods (`pct_upreg_sarc`), traditional methods (`pct_upreg_trad`), and no method (`pct_upreg_nouse`). No method variables can serve as control groups for statistical tests. Covariates will include the number of women using each of these methods across four different age groups: 15-19, 20-24, 25-34, and 35-49. 
-Number of women 15-19 using long-acting reversible methods `larc_1519`
-Number of women 15-19 using short-acting methods `sarc_1519`
-Number of women 15-19 using traditional methods `tradmethods_1519`
-Number of women 15-19 using no method `nomethod_1519`
-Number of women 20-24 using long-acting reversible methods `larc_2024`
-Number of women 20-24 using short-acting methods	`sarc_2024`
-Number of women 20-24 using traditional methods	`tradmethods_2024`
-Number of women 20-24 using no method	`nomethod_2024`
-Number of women 25-34 using long-acting reversible methods	`larc_2534`
-Number of women 25-34 using short-acting methods	`sarc_2534`
-Number of women 25-34 using traditional methods	`tradmethods_2534`
-Number of women 25-34 using no method	`nomethod_2534`
-Number of women 35-49 using long-acting reversible method	`larc_3549`
-Number of women 35-49 using short-acting method	`sarc_3549`
-Number of women 35-49 using all traditional methods	`tradmethods_3549`
-Number of women 35-49 using no method	`nomethod_3549`

code for low, medium, and high use per country using case_when()

```{r}
#| message: false
# select the outcome variable, predictor variables, and covariates  
rawdata2 <- rawdata %>% 
                select(country, region, subregion, pct_upreg, upreg_subreg, pct_upreg_larcster, pct_upreg_sarc, pct_upreg_trad, pct_upreg_nouse, larc_1519, sarc_1519, tradmethods_1519, nomethod_1519, larc_2024, sarc_2024, tradmethods_2024, nomethod_2024, larc_2534,  sarc_2534, tradmethods_2534,  nomethod_2534, larc_3549, sarc_3549, tradmethods_3549, nomethod_3549)

# take a glimpse at the selected variables 
skim(rawdata2)
head(rawdata2)
```

@@

First, the `region` variable is a character variable, but it will be more useful later as a factor. We can easily mutate it from a character variable to factor variable. All the remaining predictor and covariate variables are numeric as they should be. 
```{r}
#| message: false
# convert character variable to factor variable
rawdata2 <- rawdata2 %>% 
              mutate(region = as.factor(region))

@@

Second, there are 24 entries missing for the percentage of all unintended pregnancies by country `pct_upreg`, which accounts for 18.18% of the observations. After isolating the observations with this missing variable, we see that the majority of countries in this category are located in Asia. This does not make sense because the majority of countries in the raw dataset are located in Africa, which suggests that the data are not missing at random. 
```{r}
#| message: false
# find missing percentage for each variable
miss_var_summary(rawdata2)

@@

We should impute the missing data here to avoid introducing bias because 18.18% of the data is missing, and this variable is essential for answering the research question. After finding the summary statistics of this variable, the median is 0.4701 and the mean is 0.4679, which suggests that there are minimal outliers in the data.  
```{r}
#| message: false
# find mean to use for imputed data
summary(rawdata2$pct_upreg)

@@

```
Once the missing values have been imputed, we should further clean the `pct_upreg` variable to have a typical percentage format. 
```{r}
#| message: false
# convert percentage variable to typical percentage format
rawdata2 <- rawdata2 %>% 
              mutate(pct_upreg = pct_upreg * 100)

# ensure mutation worked correctly
summary(rawdata2$pct_upreg)
```

We also need to evaluate the 108 missing entries of the percentage of unintended pregnancies by subregion `upreg_subreg`. Since 81.81% of this variable is missing, it does not make sense to include it in the analysis because we do not have enough information to impute the missing data. We will remove this variable for the sake of this analysis.
```{r}
#| message: false
# isolate the percentage of all unintended pregnancies by subregion to discover trends in missingness
variable_with_missing <- "upreg_subreg"
missing_obs2 <- rawdata2[is.na(rawdata2[[variable_with_missing]]), ]
other_variables2 <- colnames(rawdata2)[!colnames(rawdata2) == variable_with_missing]
pct_subregmissing <- missing_obs2[, other_variables2]
print(pct_subregmissing)

@@

Fortunately, the rest of the variables required for this analysis are complete. Each of the four variables describing the percentage of unintended pregnancies based on birth control type also need to be converted to traditional percentage format. 
```{r}
#| message: false
# convert percentage variable to typical percentage format
rawdata2 <- rawdata2 %>% 
              mutate(pct_upreg_larcster = pct_upreg_larcster * 100, 
                     pct_upreg_sarc = pct_upreg_sarc * 100, 
                     pct_upreg_trad = pct_upreg_trad * 100, 
                     pct_upreg_nouse = pct_upreg_nouse * 100)

# ensure mutations worked correctly
summary(rawdata2$pct_upreg_larcster)
summary(rawdata2$pct_upreg_sarc)
summary(rawdata2$pct_upreg_trad)
summary(rawdata2$pct_upreg_nouse)
```

To make the covariate variables more useful, it makes sense to create a new factor variable denoting low, medium, and high usage of each birth control method by country. First, we need to determine the boundaries for each level of usage by exploring how many women use each method across age groups. We need to create a variable showing the total usage of each birth control method to remove the effect of age.  
```{r}
#| message: false
# create new variables showing total usage of each method including all age groups
rawdata2 <- rawdata2 %>% 
              mutate(total_larc = larc_1519 + larc_2024 + larc_2534 + larc_3549,
                     total_sarc = sarc_1519 + sarc_2024 + sarc_2534 + sarc_3549,
                     total_trad = tradmethods_1519 + tradmethods_2024 + tradmethods_2534 + tradmethods_3549,
                     total_none = nomethod_1519 + nomethod_2024 + nomethod_2534 + nomethod_3549)

# explore summary of each total to set boundaries for each level of usage
rawdata2 %>% skim(total_larc, total_sarc, total_trad, total_none)
```
After finding the distribution of number of women using each type of birth control, we can use that information to create a new factor variable denoting low, medium, and high usage for each method. The "low" category will include observations below the 25th percentile. The "medium" category will include observations between the 25th and 75th percentiles. The "high" category will include observations above the 75th percentile. 
```{r}
#| message: false
# create usage variable for long acting reversible methods 
rawdata2 <- rawdata2 %>% 
              mutate(larc_usage = as.factor(case_when(
                                  total_larc < 11.5 ~ "Low",
                                  total_larc >= 11.5 & total_larc < 407.4633 ~ "Medium",
                                  total_larc >= 407.4633 ~ "High")))

# check if new variable was created correctly 
class(rawdata2$larc_usage)

# create usage variable for short acting reversible methods 
rawdata2 <- rawdata2 %>% 
              mutate(sarc_usage = as.factor(case_when(
                                  total_sarc < 114.33113 ~ "Low",
                                  total_sarc >= 114.33113 & total_larc < 1554.7609 ~ "Medium",
                                  total_sarc >= 1554.7609	 ~ "High")))

# check if new variable was created correctly 
class(rawdata2$sarc_usage)

# create usage variable for traditional methods 
rawdata2 <- rawdata2 %>% 
              mutate(trad_usage = as.factor(case_when(
                                  total_trad < 15.05000	 ~ "Low",
                                  total_trad >= 15.05000 & total_larc < 332.4750 ~ "Medium",
                                  total_trad >= 332.4750	 ~ "High")))

# check if new variable was created correctly 
class(rawdata2$trad_usage)

# create usage variable for no methods
rawdata2 <- rawdata2 %>% 
              mutate(none_usage = as.factor(case_when(
                                  total_none < 80.07500	 ~ "Low",
                                  total_none >= 80.07500 & total_larc < 875.2750 ~ "Medium",
                                  total_none >= 875.2750	 ~ "High")))

# check if new variable was created correctly 
class(rawdata2$none_usage)
```

@@

```{r}
#| message: false
processeddata <- rawdata2
```

@@

```{r}
#| message: false
save_data_location <- here("data","processed-data","processeddata.rds")
saveRDS(processeddata, file = save_data_location)
```

### 2024-02-18 products/manuscript/Manuscript.qmd
Part 2 Updates
I found new variables to use as covariates for our updated research question. I added those variables to the raw data that is being processed, explored them, and created a new variable. I also made notes on how to start exploring the data.

# Potential Research Questions
1) What is the relationship between rates of unintended pregnancies and usage of contraceptive methods?
  -outcome variables: 
  unintended pregnancies, total number (pct_upreg) 
  unintended pregnancies, total number [subregional]      (upreg_subreg)

  -predictor variables:
  number of women 35-49 using modern methods (modmmethods_3549)
  number of women 35-49 using sterilization (ster_3549)
  number of women 35-49 using long-acting reversible methods (larc_3549)
  number of women 35-49 using short-acting methods (sarc_3549)
  number of women 35-49 using traditional methods (tradmethods_3549)
  number of women 35-49 using no method (nomethod_3549)

We will be looking for patterns in contraceptive methods and higher unintended pregnancies to investigate the effectiveness of each method. We hope to find relationships between subregions, rates of unintended pregnancies, and contraceptive methods to determine possible interventions that will increase the effectiveness of contraceptive methods in each subregion. Each of the predictor variables exist for two other age ranges (20-24 and 25-34), so we could analyze contraceptive usage across age categories within each subregion as well.

>>>>

# Research Questions
1) What is the relationship between rates of unintended pregnancies and usage of contraceptive methods across various age groups?
The main outcome variables will be total percentage of unintended pregnancies (`pct_upreg`). 
The predictor variables will include percentage of unintended pregnancies across three types of birth control: long acting reversible methods (`pct_upreg_larcster`), short term reversible methods (`pct_upreg_sarc`), traditional methods (`pct_upreg_trad`), and no method (`pct_upreg_nouse`). No method variables can serve as control groups for statistical tests.
Covariates will include the number of women using each of these methods across four different age groups: 15-19, 20-24, 25-34, and 35-49. 
-Number of women 15-19 using long-acting reversible methods `larc_1519`
-Number of women 15-19 using short-acting methods `sarc_1519`
-Number of women 15-19 using traditional methods `tradmethods_1519`
-Number of women 15-19 using no method `nomethod_1519`
-Number of women 20-24 using long-acting reversible methods `larc_2024`
-Number of women 20-24 using short-acting methods	`sarc_2024`
-Number of women 20-24 using traditional methods	`tradmethods_2024`
-Number of women 20-24 using no method	`nomethod_2024`
-Number of women 25-34 using long-acting reversible methods	`larc_2534`
-Number of women 25-34 using short-acting methods	`sarc_2534`
-Number of women 25-34 using traditional methods	`tradmethods_2534`
-Number of women 25-34 using no method	`nomethod_2534`
-Number of women 35-49 using long-acting reversible method	`larc_3549`
-Number of women 35-49 using short-acting method	`sarc_3549`
-Number of women 35-49 using all traditional methods	`tradmethods_3549`
-Number of women 35-49 using no method	`nomethod_3549`

We will be looking for patterns in contraceptive methods and higher unintended pregnancies across age groups to investigate the effectiveness of each method. We hope to find relationships between subregions, rates of unintended pregnancies, and contraceptive methods to determine possible interventions that will increase the effectiveness of contraceptive methods.
